---
title: "Starting with with the dplyr, ggplot2 packages, and the pipe operator |>"
author: "Han Olff - chaged by Leonor"
date: "2024-08-29"
format: html
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

# Working with the dplyr and ggplot2 packages

Load the required libraries :

```{r message = FALSE, warning = FALSE}
# clear the R environment

renv::restore()
# restore your library (keep at the start of every script)

rm(list = ls())

# load the required packages
library(tidyverse)  # this loads a group of related packages, including readr, dplyr and ggplot2
```

We will work with an online database of the Schiermonnikoog transect
study, that you will collect additional data for next week in the field.
In this example, we work with measurements of cockles (a bivalve
mollusc) on their width and length. From the database Epibenthos, we
will work with the table
[FactCockles](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1538766002#gid=1538766002).
See the documentation of the different variables in the table
[MetVariables](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1290622213#gid=1290622213)

### Read the datafile from the Google sheets database

To read the data in R, you first need to know the published csv link of
the FactCockles table. In the database, you can find this link in the
table MetTables. It is produced in Google Sheets throught the menu
File/Share/Publish to web and then selecting the table and output as
csv.

Read the FactCockleSize table with read_csv, which reads it as tibble
(formatted dataframe)

```{r}
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv")
print(FactCockleSize)
view(FactCockleSize)
names(FactCockleSize) #variables
nrow(FactCockleSize)
```

Plot with ggplot the relation between cockle thickness (thickness_mm, as
x) and length (length_mm), showing each year with a different color, and
add a regression line through all the years

```{r}
ggplot2::ggplot(data = FactCockleSize,
                mapping = aes(x=length_mm, y=thickness_mm)) +
              geom_point()
#ggplot with points


#there is a big outlier, likely a data mistake, we want to REMOVE IT
FactCockleSize |> dplyr::filter(thickness_mm>500)
#we can see that is the observation 1531

#it is handy to create the least possible data sets, so we will remove the outlier directly from the source data
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv") |> dplyr::filter(CockleObs_ID!=1531)

#but we still see some outlier in thickness_mm>10 & length_mm<5
FactCockleSize |> dplyr::filter(thickness_mm>10 & length_mm<5)
#we can see that is observation 469

#so we also remove it from the source
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv") |> dplyr::filter(CockleObs_ID!=1531, CockleObs_ID!=469)

#and we look again to the plot
ggplot2::ggplot(data = FactCockleSize,
                mapping = aes(x=length_mm, y=thickness_mm)) +
              geom_point()

```

Further explore the plot with a regression line through all the data

```{r}
#plot data with regression line
ggplot2::ggplot(data = FactCockleSize,
                mapping = aes(x=length_mm, y=thickness_mm)) +
              geom_point() + 
              geom_smooth(method = "lm")

#calculate a linear regression
model_lm <- lm(thickness_mm~length_mm, data= FactCockleSize)
# linear model means that y is explain (~) by x, in my specific data
summary(model_lm)
#Coeficient ''Estimate'' results
#equation: thickness_mm= -0.353 + 0699xlength_mm

#a cockle of length 20mm has a predicted with of:
-0.353 + 0.699*20 #13.62

#also this regression is significant R2=0.899, P-value<0,0001

```

Make same plot but showing a separate regression line per year

```{r}
# color the points by year, but plot one regression line

#turn color as factor
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv") |> dplyr::filter(CockleObs_ID!=1531, CockleObs_ID!=469) |> # USE PIPE!!
dplyr::mutate(year=factor(year)) # make year into a factor

#assign colour into your observation depending in the year ()
ggplot2::ggplot(data = FactCockleSize,
                mapping = aes(x=length_mm, y=thickness_mm)) +
              geom_point(mapping = aes(col=year)) + 
              geom_smooth(method = "lm")

```

Make a panel plot where with each year is shown as a separate graph

```{r}
#color the points but in diferent plots
ggplot2::ggplot(data = FactCockleSize,
                mapping = aes(x=length_mm, y=thickness_mm)) +
              geom_point(mapping = aes(col=year)) + 
              geom_smooth(method = "lm") + 
              facet_wrap(~year)

#but now exclude years without data
drop_na(FactCockleSize) |> #take out the missing data
  ggplot2::ggplot(mapping = aes(x=length_mm, y=thickness_mm)) +
              geom_point(mapping = aes(col=year), size=2) + 
              xlab("cockle length") +
              ylab("cockle thickeness") +
              geom_smooth(method = "lm") +
              facet_wrap(~year)
# add labels
#xlab("cockle length") +
#ylab("cockle thickeness)

```

We conclude from this that: \* there were two important outliers in the
dataset that were removed after visual inspection \* the regression
between length and width is abou the same for every year, we safely can
use only length as a proxy for the biomass of an individual cockle
